---
title: "Creating Clade A v Clade C Balanced DF"
author: "Cassandra Sperow"
date: "2023-11-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(library(tidyverse))
```

## 
## Read in Data
```{r}
read_csv("../output/rev_coral_join.csv") -> rev_coral_join
```

- The above import of data has 778 rows because some samples have sequencing counts for more than one Clade. 


## Subset Clade C as it's the minority
- Clade A is the majority
- Clade D, G and B only have 1 or 2 rows; therefore, need to focus on Clade A or Clade C. 
```{r}

# Extract the minority class samples (139 observations)
minority_class_C <- rev_coral_join[rev_coral_join$Clade == 'C', ] # 139

# Extract the majority class samples (the rest of the dataset)
majority_class_A <- rev_coral_join[rev_coral_join$Clade == 'A', ] # 628

######### Percent Clade C v Clade A
139/(139 + 628)
```

- 18 % is not bad for the minority class, so since it's better to go with more data, we will keep Clade A with over 600 samples so that we can model it better. Clade C was already modeled (see 'Ridge & Lasso Modeling.Rmd') but Clade A will be modeled in python for Data 642 project. 

```{r}

# Combine the datasets
AC_df <- rbind(majority_class_A, minority_class_C)

# Shuffle the rows to randomize the order
AC_df <- AC_df[sample(nrow(AC_df)), ]

```

```{r}
dim(AC_df) # 767 x 5897
head(AC_df)
```

```{r}


```


```{r}
write_csv(AC_df, "../output/AC_df.csv")
```


```{r}
AC_df %>% 
  ggplot(aes( x = Clade, color = Clade, fill = Clade)) +
  geom_histogram(stat = "count") +
  theme_bw() +
  ggtitle("Clade A and C Class Labels")
```

# Make files with CLR transofmraiton for first 500, 1000,  2000
```{r}
# clr on first 500 ASVs
bind_cols(
  AC_df[,1:8], 
  AC_df[,9:508] %>% 
    compositions::clr() 
) -> AC_clr_df_500

# clr on first 1000
bind_cols(
  AC_df[,1:8], 
  AC_df[,9:1008] %>% 
    compositions::clr() 
) -> AC_clr_df_1000

# clr on first 2000
bind_cols(
  AC_df[,1:8], 
  AC_df[,9:2008] %>% 
    compositions::clr() 
) -> AC_clr_df_2000
```





```{r}
write_csv(AC_clr_df_500, "../output/AC_clr_df_500.csv")
write_csv(AC_clr_df_1000, "../output/AC_clr_df_1000.csv")
write_csv(AC_clr_df_2000, "../output/AC_clr_df_2000.csv")
```


```{r}

GGally::ggpairs(AC_clr_df_1000[, c(2, 9:18)], aes(color = Clade)) -> ggpairs_AC


ggsave("../plots/ggpairs_AC.png", 
       width = 11, height = 12)

ggpairs_AC
```


```{r}

# Create density plots for each numeric column
density_plots <- map(names(AC_clr_df_1000)[9:20], 
                     function(var) {
                        ggplot(AC_clr_df_1000, aes(x = AC_clr_df_1000[[var]], fill = Clade)) +
    geom_density(alpha = 0.5) +
    ggtitle(paste( var)) +
    theme_minimal() +
                         xlab(NULL) 
                        # theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())  
})


```


```{r}
density_plots
```


```{r}
# Arrange the plots on a grid
library(cowplot)
plot_grid(plotlist = density_plots, ncol = 4)  # Change 'ncol' as needed


```


```{r}
```


```{r}
```

