---
title: "Coral_Data_EDA"
author: "Cassandra Sperow"
date: "2023-10-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)

library(tidyverse)
```

## Exploratory Data Analysis of Joined Algae Clade & Bacteria in Coral Samples

**Note: Please see revised data frame as rev_coral_join.csv that prioritizes clade C.**

- Read in previously joined algae and bacteria data per coral. For the join process, see 'Join.Rmd'.  
```{r}
read_csv("../output/coral_data.csv") -> corals

#ITS2_f should be a factor
corals %>% 
  mutate(ITS2_f = as.factor(ITS2_f)) -> corals
```


```{r}
dim(corals)
```

### What are the proportions of observations per clade?
```{r}
corals %>% 
  group_by(Clade) %>% 
  summarise(n = n()) 
  
```

- Clade A has 606 observations. 606/(606+52) = about 92 % 

- Clade C has 52 observations. 52/(606+52) = about 8 %

Notes:  If we are trying to explain the variability of clades based on bacteria, logistic regression would be used to explain or predict either clade A or C with bacteria ASVs as columns; however, taking out other info (such as region) is not advisable as they are all possible predictors. 

- How correlated are the top 2 ASVs with each of the algae variables?

- Logging the ASV counts presents better visualization:

```{r}
corals %>% 
  ggplot(aes(x = log(ASV0001), y = log(ASV0002), color = Clade)) +
  geom_point(position = "jitter") + 
  facet_wrap(~species)
```

- Clade A is far more prevalent as is also shown in the number of abservations per Clade based on these data.

- Distributions are different per species. ASV1 and 2 are more clustered for species P and more spread out in terms of counts for species S. 

- From coseq() K-means clustering, this is perhaps why ASV1 and 2 were consistently clustered together in species P but not for species S. Perhaps ASVs 1 and 2 play a different role in species S. 

## Data Subsetting of first 1000 most abundant ASVs per Species
- Recommendation to reduce dimensionality:  After the coral data are split by species, client approved subsetting the first 1000 ASVs. 

- Recommendation for binary classification:  Use Principal Components Analysis (PCA) in the file 'PCA_CLR.Rmd' to obtain the most contributing ASVs per species to make sure they are in the top 1000 (or fewer if reduced again later for computational limits). 


### Species P Data Subsetting - Which ASVs are not found in at least 3 samples for Species P?
```{r}
corals %>% 
  filter(species =="P") %>%  # 296:  cut off will be 293 inclusive
  select(-c(1:8)) -> corals_p # saving in order to pass to map() function

### These are the ASVs that are found in at least 3 samples or more for species P
map(corals_p, ~sum(.==0)) %>% 
  as_tibble() %>% 
  pivot_longer(cols = everything(), 
               names_to = "ASV_column", 
               values_to = "sum_col_eq_0") %>% 
  arrange(sum_col_eq_0) %>% 
  # cut off includes 293 or more because 296-293 = 3 
  filter(sum_col_eq_0 >=293) -> asvs_speciesP
```

- Which ASVs are NOT found in Species P?  ASV3
```{r}
map(corals_p, ~sum(.==0)) %>% 
  as_tibble() %>% 
  pivot_longer(cols = everything(), 
               names_to = "ASV_column", 
               values_to = "sum_col_eq_0") %>% 
  # cut off includes 293 or more because 296-293 = 3 
  filter(sum_col_eq_0 == 0)
```

- Take out ASV3 for species P data subset
```{r}
# use original data frame to get other columns back into data subset
corals %>% 
  filter(species == "P") %>% 
  # take out asv 3 bc we know it's not found in this species
  # take out species column 
  select(-ASV0003, -species) %>% 
  group_by(Clade  ) %>% 
  summarise(n = n())
```

- Note: 13/283 is about 5 % 

### Species S Data Subsetting - Which ASVs are not found in at least 3 samples for Species S?

```{r}
corals %>% 
  filter(species =="S") %>%  # 362:  cut off will be 362-3 =  359 inclusive
  select(-c(1:8)) -> corals_s # saving here in order to pass to map() function

### These are the ASVs that are found in at least 3 samples or more for species S
map(corals_s, ~sum(.==0)) %>% 
  as_tibble() %>% 
  pivot_longer(cols = everything(), 
               names_to = "ASV_column", 
               values_to = "sum_col_eq_0") %>% 
  arrange(sum_col_eq_0) %>% 
  # cut off includes 293 or more because 296-293 = 3 
  filter(sum_col_eq_0 >=359) -> asvs_speciesS

```

- Which ASVs are NOT found in Species S ?  All 5,889 ASVs are found in Species S. 
```{r}
map(corals_s, ~sum(.==0)) %>% 
  as_tibble() %>% 
  pivot_longer(cols = everything(), 
               names_to = "ASV_column", 
               values_to = "sum_col_eq_0") %>% 
  # cut off includes 359 or more 
  filter(sum_col_eq_0 == 0)

dim(corals_s)
head(corals_s)
```


### Visualization Experiment with t-SNE Plot: 
- t-SNE is a way to visualize high-dimensional data and is recommended for hundreds of thousands of ASVs, recommended by Teaching Assistant for Data 793. 
- On a separate occassion previously, Client indicated that it is also possible to use first 1000 ASVs as these are already ranked in terms of the abundance they were found in the data. 

Code adapted from [here](https://plotly.com/r/t-sne-and-umap-projections/)
```{r}
library(Rtsne)
```

- For t-SNE, select top 1000 ASVs no matter species to test out visualization
```{r}
corals %>% 
  select(starts_with("ASV")[1:1000]) %>% 
  as.matrix() -> asv_matrix
```


### Example t-SNE visualization

```{r}
library(tsne)
library(plotly)
data("iris")

features <- subset(iris, select = -c(Species)) 

set.seed(0)
tsne <- tsne(features, initial_dims = 2)
tsne <- data.frame(tsne)
pdb <- cbind(tsne,iris$Species)
options(warn = -1)
fig <-  plot_ly(data = pdb ,x =  ~X1, y = ~X2, type = 'scatter', mode = 'markers', split = ~iris$Species)

fig <- fig %>%
  layout(
    plot_bgcolor = "#e5ecf6"
  )

fig

```

```{r}
dim(asv_matrix)
```

## t-SNE on coral samples and top 1000 ASVs - Both Species
```{r, cache=TRUE}
library(tsne)
library(plotly)


features <- subset(asv_matrix) # 658 x 1000

set.seed(0)
tsne <- tsne(features, initial_dims = 2)
tsne <- data.frame(tsne)
pdb <- cbind(tsne,corals$species)
options(warn = -1)
fig <-  plot_ly(data = pdb ,x =  ~X1, y = ~X2, type = 'scatter', mode = 'markers', split = ~corals$species)

fig <- fig %>%
  layout(
    plot_bgcolor = "#e5ecf6"
  )

fig
```


## t-SNE for Clade - Both Species - Top 1000 ASVs
```{r, cache=TRUE}
library(tsne)
library(plotly)

features <- subset(asv_matrix) 

set.seed(0)
tsne <- tsne(features, initial_dims = 2)
tsne <- data.frame(tsne)
pdb <- cbind(tsne,corals$Clade)
options(warn = -1)
fig2 <-  plot_ly(data = pdb ,x =  ~X1, y = ~X2, type = 'scatter', mode = 'markers', split = ~corals$Clade)

fig2 <- fig2 %>%
  layout(
    plot_bgcolor = "#e5ecf6"
  )

fig2
```

## t-SNE for Species P
- This took excessively long to run and had to be stopped, but it had the 5,889 ASVs
- Try again with subset of the Top 1000 ASVs: 
```{r, cache = TRUE}
# library(tsne)
# library(plotly)
# 
# 
# 
# features <- as.matrix(corals_p, select..........)
# 
# set.seed(0)
# tsne <- tsne(features, initial_dims = 2)
# tsne <- data.frame(tsne)
# pdb <- cbind(tsne,corals$Clade)
# options(warn = -1)
# fig <-  plot_ly(data = pdb ,x =  ~X1, y = ~X2, type = 'scatter', mode = 'markers', split = ~corals$Clade)
# 
# fig <- fig %>%
#   layout(
#     plot_bgcolor = "#e5ecf6"
#   )
# 
# fig
```

## GGpairs on top 10 ASVs and Clade for Species P
```{r, cache=TRUE}

corals[, -c(1,3,4,5,7,8) ] %>% 
  filter(species=="P") %>% 
  select(1, 3:12) %>% 
  GGally::ggpairs(aes(color = Clade)) -> ggpairs_species_P

ggpairs_species_P

```

```{r}
ggsave( "../plots/ggpairs_species_P.png", ggpairs_species_P, width = 14)
```

## GGpairs on top 10 ASVs and Clade for Species P
```{r, cache=TRUE}

corals[, -c(1,3,4,5,7,8) ] %>% 
  filter(species=="S") %>% 
  select(1, 3:12) %>% 
  GGally::ggpairs(aes(color = Clade)) -> ggpairs_species_S

ggpairs_species_S
```

```{r}
ggsave( "../plots/ggpairs_species_S.png", ggpairs_species_S, width = 14)

```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```

