---
title: "Correlation Plots & Clustering Analysis with Ridge & Lasso Predictors"
author: "Cassandra Sperow"
date: "2023-11-12"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(library(tidyverse))
```


- Note: The first several correlation plots relate to exploratory data analysis while other sections towards the end relate to post-modeling analysis with the predictors from 2 rounds of Ridge, Lasso, and tree methods modeling (See Rmd files with Ridge, Lasso, and/or tree methods in title). The Ridge & Lasso Modeling Revised.Rmd file is the final results of modeling Clade C vs. non-Clade C. 

# Clade C

```{r}
read_csv("../output/lr_data.csv") %>% 
  mutate(Clade = as.factor(Clade)) -> lr_data

head(lr_data)
```

# Correlation Matrix - Clade C - All
```{r}
cor_matrix_lr_data <- cor(lr_data[,-c(1,2)]) # 3226

```


## Correlation Matrix - Top 20
```{r}



cor_matrix_lr_data[1:20,1:20]
```

```{r}
png(filename = "../plots/top20_corr_clade_c.png",
    width = 1200, height = 800)

library(corrplot)
top20_corr_clade_c <- corrplot(cor_matrix_lr_data[1:20, 1:20], 
                               method = "circle")

```



## Correlation Matrix - Top 50
- Anything over 50 gets small and hard to read
```{r}
png(filename = "../plots/top50_corr_clade_c.png",
    width = 1200, height = 800)

library(corrplot)
top50_corr_clade_c <- corrplot(cor_matrix_lr_data[1:50, 1:50], 
                               method = "circle")

```

- preliminary
```{r}
# read in taxonomy table to see which bacterium this is
read_delim("../data/Files/16S_tax.txt") -> bac_tax

# one of the earlier experimental models had ASV0013 kept as predictor
bac_tax %>% 
  filter(ASV == "ASV0013") %>% 
  pivot_longer(cols = everything())
```

- preliminary
```{r}
#plot(log(lr_data$ASV0013), lr_data$Clade)

ggplot(lr_data, aes( y = log(ASV0013), color = as.factor(Clade), fill = as.factor(Clade))) +
  geom_histogram() +
  coord_flip() +
  facet_wrap(~Clade) +
  xlab("Log of ASV0013 Counts") +
  ylab("")
```

# Correlation Plot - Both Species - All Data
```{r}
read_csv("../output/rev_coral_join.csv") -> rev_coral_join
```


```{r}
# correlation matrix
corr_matrix_all <- cor(rev_coral_join[,-c(1:8)])

corr_matrix_all[1:20,1:20]

# save the png
png(filename = "../plots/corr_plot_top_20.png",
    width = 1200, height = 900)

# correlation plot
library(corrplot)
corr_plot_top_20 <- corrplot(corr_matrix_all[1:20, 1:20], 
                               method = "circle", 
                             )

#corr_plot_top_20
```


```{r}
# save the png
png(filename = "../plots/corr_plot_top_100.png",
    width = 1200, height = 900)

# correlation plot
library(corrplot)
corr_plot_top_100 <- corrplot(corr_matrix_all[1:100, 1:100], 
                               method = "circle", 
                              )
#corr_plot_top_100
```


```{r}
# save the png
png(filename = "../plots/corr_plot_top_50.png",
    width = 1200, height = 900)

# correlation plot
library(corrplot)
corr_plot_top_50 <- corrplot(corr_matrix_all[1:50, 1:50], 
                               method = "circle", 
                              )

corr_plot_top_50
```

----Ridge modeling coefficients for 'best' model - first round of modeling - not final
- Reader is advised to skip to section below of: **Revised/Final Modeling Results**

- See the below section for the final results for Ridge 200 in 2nd (final) round of modeling with lower error rates
```{r}
read_csv("../output/ridge_500_coef_df.csv") -> ridge_results
```

```{r}
ridge_results
```
-Where are these ASVs in relation to one another? - See revised final models below 

```{r}
cor_matrix_lr_data %>% # data used in clade C modeling
  as.data.frame() %>% 
  mutate(ASV = rownames(.), .before = ASV1236) %>% 
 # filter(ASV %in% c(ridge_results$row_names)) %>% 
  filter(ASV %in% c("ASV6296"	,	"ASV4003"	,"ASV8219"	,	"ASV4449"	) )
  
```
- They are uncorrelated, and ridge would have chosen these because of this.

---Lasso modeling selection - round 1 - not final
```{r}
read_csv("../output/lasso_500_coef_df.csv") -> lasso_results

lasso_results

cor_matrix_lr_data %>% # data used in clade C modeling
  as.data.frame() %>% 
  mutate(ASV = rownames(.), .before = ASV1236) %>% 
  filter(ASV %in% c("ASV0675"	,	"ASV0013"	,	"ASV0958"	,	"ASV0485"))
```
--Were these clustered consistently in the k-means runs?
```{r, read-in-cluster-assignments}
read_csv("../output/species_P_cluster_assignments.csv") -> clusters_P
read_csv("../output/species_S_cluster_assignments.csv") -> clusters_S

```

```{r}
dim(clusters_P)
dim(clusters_S)
dim(ridge_results)
```

```{r}
ridge_results %>% 
  left_join(clusters_P, join_by("row_names" == "ASV"))


# what is biggest group?
ridge_results %>% 
  left_join(clusters_P, join_by("row_names" == "ASV")) %>% # cluster 3 under seed23_run2
  group_by(seed12_run4) %>% 
  summarise(n = n()) %>% # cluster 3 in seed 23just happens to be largest group
  arrange(-n) # cluster 1 in seed 12 - same though with jaccard index
```

- The ridge results above joined with the clustering results for species P indicate that the first several ASVs with the highest absolute value coefficient in ridge modeling were clustered together consistently in predominantly 3 out of 3 random runs. 

- What does the same thing look like for species S?
```{r}
ridge_results %>% 
  left_join(clusters_S, join_by("row_names" == "ASV"))


ridge_results %>% 
  left_join(clusters_S, join_by("row_names" == "ASV")) %>% # cluster 10 under seed23_run2
  group_by(seed12_run4) %>% 
  summarise(n = n()) %>% # cluster 10 in seeed 23
  arrange(-n) # cluster 6 in seed 12 but same cluster
```

- Yes, the ASVs that the ridge model has as its highest absolute value are clustered pretty consistently across 3 out of 3 random runs of K-means. See Coseq_S and Coseq_P files in analysis folder for the k-means runs per species. 

# **Revised/Final Modeling Results**
- Skip to Recommendation below
- The above was from the first rounds of modeling efforts that never got better than around a 30 % error rate. 
- This section uses the revised (final) modeling results where the 'best' model found on these data is the Ridge 200 with 200 ASV predictors using also other non-ASV predictors.
- See Ridge and Lasso Modeling Revised.Rmd in analysis folder. 

```{r}
readRDS("../output/lasso_coefs.rds") -> lasso_coefs
readRDS("../output/ridge_coefs.rds") -> ridge_coefs
```

```{r, eda-0}

lasso_coefs[3] %>%  #%>% #
  as.data.frame() %>%
  # exclude other variables except bacteria
  filter(str_detect(predictor, "ASV")) %>%
  # order by largest magnitude
  arrange(-abs(s1)) %>%
  select(predictor) %>%
  as_vector() -> lasso_1000_best_asvs
  
```

```{r, eda-1}
dim(cor_matrix_lr_data) # all Clade C 3226
cor_matrix_lr_data %>% # data used in clade C modeling
  as.data.frame() %>% 
  mutate(ASV = rownames(.), .before = ASV1236) %>% 
 # match the asv above
  filter(ASV %in% c( lasso_1000_best_asvs))
```


### Join Predictors with Cluster Assignments & Review Consistency of Cluster Assignments

```{r, eda-2}
lasso_coefs[3] %>% # lasso 1000
  as.data.frame() %>% 
  filter(str_detect(predictor, "ASV")) %>% 
  left_join(clusters_P, join_by("predictor" == "ASV")) %>% 
  head




lasso_coefs[3]  %>% # lasso 1000
  as.data.frame() %>% 
  filter(str_detect(predictor, "ASV")) %>% 
  left_join(clusters_S, join_by("predictor" == "ASV")) %>% 
  head
```



### Join Predictors with Cluster Assignments for Largest Cluster
- Use Seed 12 because it had highest Jaccard Index
```{r, eda-3}
lasso_coefs[3] %>% # lasso 1000
  as.data.frame() %>% 
  filter(str_detect(predictor, "ASV")) %>% 
  left_join(clusters_P, join_by("predictor" == "ASV")) %>% 
  group_by(seed12_run4) %>% 
  summarise(n = n()) %>% # mostly cluster 3
  arrange(-n)




lasso_coefs[3]  %>% 
  as.data.frame() %>% 
  filter(str_detect(predictor, "ASV")) %>% 
  left_join(clusters_S, join_by("predictor" == "ASV")) %>% 
  group_by(seed12_run4) %>% 
  summarise(n = n()) %>% # mostly cluster 10
  arrange(-n)
```


# Read in Bacteria Taxonomy
```{r, bac-taxonomy}
read_delim("../data/Files/16S_tax.txt") -> bac_tax
```

# Join Predictors with Cluster Assignments & Taxonomy to Review 
- Cluster runs with K-means were completed by species
- See also 'coseq' files in analysis folder
- Recommended model is Ridge 200; therefore, lasso 1000 plots are commented out.

# Ridge 200 - **Recommended Model Overall**
- comparable low error rate relative to all other modeling methods
- higher interpretability with keeping predictors (unlike Lasso)
- less complex
- less expensive computationally

```{r, joins-ridge-200}
################### combine ridge 200 predictors with cluster assignments and taxonomy
# species P
ridge_coefs[3] %>% # Ridge 200
  as.data.frame() %>% 
  filter(str_detect(predictor, "ASV")) %>% 
  left_join(clusters_P, join_by("predictor" == "ASV")) %>% 
  left_join(bac_tax, join_by("predictor" == "ASV")) %>% 
  group_by(Family, Genus) %>%  # 16 groups
  ungroup() %>% 
  mutate(coral_species = "P") -> predictors_p_df

# species S
ridge_coefs[3]  %>% 
  as.data.frame() %>% 
  filter(str_detect(predictor, "ASV")) %>% 
  left_join(clusters_S, join_by("predictor" == "ASV")) %>% 
  left_join(bac_tax, join_by("predictor" == "ASV")) %>% 
  group_by(Family, Genus) %>%  # 16 groups
  ungroup() %>% 
  mutate(coral_species = "S") -> predictors_s_df

# combine
bind_rows(
predictors_p_df, 
predictors_s_df) -> predictors_clade_c_ridge # genus or family are best per client.


predictors_clade_c_ridge %>% # 400 because the k-means was done for each species separately
  arrange(-abs(s1)) # in order of coefficient absolute value: high to low
```

- The interpretation of the coefficients from a ridge model with a bias term is that the magnitude above the approach to 0 may indicate greater influence over the outcome of Clade C or not Clade C. 



```{r, plots-ridge-200}

# predictor cluster assignments from ridge 200
# facet wrap for coral species
predictors_clade_c_ridge %>% 
  ggplot(aes(x = seed12_run4, fill = Family)) +
  geom_histogram(stat = 'count', show.legend = FALSE) +
  viridis::scale_fill_viridis(discrete = T) +
  theme_bw() +
  xlab("Cluster Assignment in Seed 12 K-Means") + 
  ggtitle("Cluster Assignnments of Ridge 200 Predictors with Taxonomy: Family") +
  facet_wrap(~paste("Species ", coral_species)) 
# facet wrap for coral species
predictors_clade_c_ridge %>% 
  ggplot(aes(x = seed12_run4, fill = Genus)) +
  geom_histogram(stat = 'count', show.legend = FALSE) +
  viridis::scale_fill_viridis(discrete = T) +
  theme_bw() +
  xlab("Cluster Assignment in Seed 12 K-Means") + 
  ggtitle("Cluster Assignnments of Ridge 200 Predictors with Taxonomy: Genus") +
  facet_wrap(~paste("Species ", coral_species)) 



################### NO facet wrap - BOTH species together
predictors_clade_c_ridge %>% 
  ggplot(aes(x = seed12_run4, fill = Family)) +
  geom_histogram(stat = 'count', 
                 #show.legend = FALSE
                 ) +
  viridis::scale_fill_viridis(discrete = T) +
  theme_bw() +
  xlab("Cluster Assignment in Seed 12 K-Means") + 
  ggtitle("Cluster Assignnments of Ridge 200 Predictors with Taxonomy: Family") +
  theme(legend.key=element_blank(), legend.key.size=unit(5,"point")) -> plot_ridge_200_family
  #guides(colour=guide_legend(nrow=10)) 
  

ggsave(filename = "../plots/plot_ridge_200_family.png",
       plot = plot_ridge_200_family, 
       device = "png",
       width = 17,
       height = 11,
       units = "in",
       )

################### facet wrap for species
predictors_clade_c_ridge %>% 
  ggplot(aes(x = seed12_run4, fill = Family)) +
  geom_histogram(stat = 'count', 
                 show.legend = FALSE
                 ) +
  viridis::scale_fill_viridis(discrete = T) +
  theme_bw() +
  xlab("Cluster Assignment in Seed 12 K-Means") + 
  ggtitle("Cluster Assignnments of Ridge 200 Predictors with Taxonomy: Family") +
  theme(legend.key=element_blank(), legend.key.size=unit(5,"point")) +
  facet_wrap(~paste("Species ", coral_species)) 


# 
# predictors_clade_c_ridge %>% 
#   ggplot(aes(x = seed23_run2, fill = Genus)) +
#   geom_histogram(stat = 'count', show.legend = FALSE) +
#   viridis::scale_fill_viridis(discrete = T) +
#   theme_bw() +
#   xlab("Cluster Assignment in Seed 23 K-Means") + 
#   ggtitle("Cluster Assignnments of Ridge 200 Predictors with Taxonomy: Genus") 

###################  facet wrap 
predictors_clade_c_ridge %>% 
  ggplot(aes(x = seed12_run4, fill = Order)) +
  geom_histogram(stat = 'count', 
                 #show.legend = FALSE
                 ) +
  viridis::scale_fill_viridis(discrete = T) +
  theme_bw() +
  xlab("Cluster Assignment in Seed 12 K-Means") + 
  ggtitle("Cluster Assignnments of Ridge 200 Predictors with Taxonomy: Order") +
  theme(legend.key=element_blank(), legend.key.size=unit(5,"point")) +
  guides(colour=guide_legend(nrow=30)) +
  facet_wrap(~paste("Species ", coral_species)) 
```



**Observations of Ridge 200 ASVs with Cluster Assignments:**

- Clustering is very consistent across runs (see also Jaccard Index calculation in 'Jaccard.Rmd'.)
- Most of the predictors are in one majority cluster assignment, e.g. first 7 are always together. 
- What does this look like in terms of Family, Genus and other taxonomy?
- Client would like to use Genus level and Family level plots in final report. 
- Recommendation is to use the Ridge 200 model because it keeps the ASVs as client would need for interpretation. Lasso cancels out most ASVs by setting them equal to 0. 



```{r, lasso-joins}
# species P
lasso_coefs[3] %>% # lasso 1000
  as.data.frame() %>% 
  filter(str_detect(predictor, "ASV")) %>% 
  left_join(clusters_P, join_by("predictor" == "ASV")) %>% 
  left_join(bac_tax, join_by("predictor" == "ASV")) %>% 
  group_by(Family, Genus) %>%  # 16 groups
  ungroup() %>% 
  mutate(coral_species = "P") -> predictors_p_df


# species S
lasso_coefs[3]  %>% 
  as.data.frame() %>% 
  filter(str_detect(predictor, "ASV")) %>% 
  left_join(clusters_S, join_by("predictor" == "ASV")) %>% 
  left_join(bac_tax, join_by("predictor" == "ASV")) %>% 
  group_by(Family, Genus) %>%  # 16 groups
  ungroup() %>% 
  mutate(coral_species = "S") -> predictors_s_df


# combine
bind_rows(
predictors_p_df, 
predictors_s_df) -> predictors_clade_c_lasso # genus or family are best per client.

```


```{r, plots-lasso-positive-negative}
predictors_clade_c_lasso %>% 
  arrange(-s1) %>% # highest positive to lowest negative magnitude
# how many are positive coefficients?
  filter(s1 > 0) -> positive_lasso

# positive_lasso
# 
# # begin plots for positive coefficients
# positive_lasso %>% 
#   ggplot(aes(x = seed12_run4, fill = Family)) +
#   geom_histogram(stat = 'count') +
#   viridis::scale_fill_viridis(discrete = T) +
#   theme_bw() +
#   xlab("Cluster Assignment in Seed 12 K-Means") + 
#   ggtitle("Cluster Assignnments of Lasso 1000 \nPositively Associated Predictors with Taxonomy") +
#   facet_wrap(~paste("Species ", coral_species)) 
# 
# positive_lasso %>% 
#   ggplot(aes(x = seed12_run4, fill = Genus)) +
#   geom_histogram(stat = 'count') +
#   viridis::scale_fill_viridis(discrete = T) +
#   theme_bw() +
#   xlab("Cluster Assignment in Seed 12 K-Means") + 
#   ggtitle("Cluster Assignnments of Lasso 1000 \nPositively Associated Predictors with Taxonomy") +
#   facet_wrap(~paste("Species ", coral_species)) 
#   
# 
# 
# 
# predictors_clade_c_lasso %>% 
#   arrange(-s1) %>% # highest positive to lowest negative magnitude
# # how many are positive coefficients?
#   filter(s1 < 0) -> negative_lasso
# 
# # begin plots for neg associated predictors
# negative_lasso %>% 
#   ggplot(aes(x = seed12_run4, fill = Genus)) +
#   geom_histogram(stat = 'count') +
#   viridis::scale_fill_viridis(discrete = T) +
#   theme_bw() +
#   xlab("Cluster Assignment in Seed 12 K-Means") + 
#   ggtitle("Cluster Assignnments of Lasso 1000 \nNegatively Associated Predictors with Taxonomy") +
#   facet_wrap(~paste("Species ", coral_species)) 
# 
# negative_lasso %>% 
#   ggplot(aes(x = seed12_run4, fill = Family)) +
#   geom_histogram(stat = 'count') +
#   viridis::scale_fill_viridis(discrete = T) +
#   theme_bw() +
#   xlab("Cluster Assignment in Seed 12 K-Means") + 
#   ggtitle("Cluster Assignnments of Lasso 1000 \nNegatively Associated Predictors with Taxonomy") +
#   facet_wrap(~paste("Species ", coral_species)) 


```

```{r, plots-lasso-1000}
# 
# predictors_clade_c_lasso %>% 
#   ggplot(aes(x = seed12_run4, fill = Phylum)) +
#   geom_histogram(stat = 'count') +
#   viridis::scale_fill_viridis(discrete = T) +
#   theme_bw() +
#   xlab("Cluster Assignment in Seed 12 K-Means") + 
#   ggtitle("Cluster Assignnments of Lasso 1000 Selected Predictors with Taxonomy") +
#   facet_wrap(~paste("Species ", coral_species)) 
# 
# predictors_clade_c_lasso %>% 
#   ggplot(aes(x = seed12_run4, fill = Class)) +
#   geom_histogram(stat = 'count') +
#   viridis::scale_fill_viridis(discrete = T) +
#   theme_bw() +
#   xlab("Cluster Assignment in Seed 12 K-Means") + 
#   ggtitle("Cluster Assignnments of Lasso 1000 Selected Predictors with Taxonomy") +
#   facet_wrap(~paste("Species ", coral_species)) 
# 
# predictors_clade_c_lasso %>% 
#   ggplot(aes(x = seed12_run4, fill = Order)) +
#   geom_histogram(stat = 'count') +
#   viridis::scale_fill_viridis(discrete = T) +
#   theme_bw() +
#   xlab("Cluster Assignment in Seed 12 K-Means") + 
#   ggtitle("Cluster Assignnments of Lasso 1000 Selected Predictors with Taxonomy") +
#   facet_wrap(~paste("Species ", coral_species)) 
# 
# predictors_clade_c_lasso %>% 
#   ggplot(aes(x = seed12_run4, fill = Family)) +
#   geom_histogram(stat = 'count') +
#   viridis::scale_fill_viridis(discrete = T) +
#   theme_bw() +
#   xlab("Cluster Assignment in Seed 12 K-Means") + 
#   ggtitle("Cluster Assignnments of Lasso 1000 Selected Predictors with Taxonomy") +
#   facet_wrap(~paste("Species ", coral_species)) 
# 
# predictors_clade_c_lasso %>% 
#   ggplot(aes(x = seed12_run4, fill = Genus)) +
#   geom_histogram(stat = 'count') +
#   viridis::scale_fill_viridis(discrete = T) +
#   theme_bw() +
#   xlab("Cluster Assignment in Seed 12 K-Means") + 
#   ggtitle("Cluster Assignnments of Lasso 1000 Selected Predictors with Taxonomy") +
#   facet_wrap(~paste("Species ", coral_species)) 
# 
# #do not use - most do not have species
# predictors_clade_c_lasso %>%
#   ggplot(aes(x = seed12_run4, fill = Species)) +
#   geom_histogram(stat = 'count') +
#   viridis::scale_fill_viridis(discrete = T) +
#   theme_bw() +
#   xlab("Cluster Assignment in Seed 12 K-Means") +
#   ggtitle("Cluster Assignnments of Lasso 1000 Selected Predictors with Taxonomy") +
#   facet_wrap(~paste("Species ", coral_species))
# 
# 
# # 
# # ggsave(plot = plot_family, "../plots/plot_family.png", 
# #        width = 12)
# # 
# # gridExtra::grid.arrange(plot_phylum, plot_class,plot_order ,plot_family, plot_genus, ncol = 3) -> five_plots
# # 
# # # Save the combined plots to a PDF file using ggsave
# # ggsave("../plots/five_plots.pdf", five_plots, width = 17, height = 11, device = "pdf")
# 
# 
# predictors_clade_c_lasso %>% 
#   ggplot(aes(x = seed12_run4, fill = Family)) +
#   geom_histogram(stat = 'count') +
#   viridis::scale_fill_viridis(discrete = T) +
#   theme_bw() +
#   xlab("Cluster Assignment in Seed 12 K-Means") + 
#   ggtitle("Cluster Assignnments of Lasso 1000 Selected Predictors with Taxonomy") 
```






