---
title: "Coseq_Species_S"
author: "Cassandra Sperow"
date: "2023-10-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(coseq)
```

## Experimental K-Means Clustering with Species S
- Using the R package 'coseq', perform K-means clustering on bacterial ASV data
```{r}
# read in the split dataset with only species S
read_csv("../output/s_bac.csv") -> s_bac
```


```{r}
dim(s_bac)
```


```{r}
head(s_bac)

s_bac[,-1] -> s_bac_num
```

## Run coseq() with K-means
- Uses MacQueen 1967 K-means algorithm 
- The below is commented out due to taking several hours to run. Results are saved in an RDS file in the output folder. 
```{r}
# set.seed(42)
# run_kmeans <- coseq(s_bac_num, # numeric only
#                     K = 2:100, 
#                     transformation = "logclr",
#                     model = "kmeans", 
#                     nstart = 100, 
#                     iter.max = 1000)
```


```{r}
#run_kmeans
```

- The first run of K-means find the "best" solution of 29 clusters.

```{r}
# clusters(run_kmeans) %>% 
#   as.data.frame() %>% 
#   `colnames<-`("cluster_29") -> clusters_29

```

- The call of the above K-means function with searching through 100 clusters with 100 nstart and 1000 iterations took several hours to run. The start time was around 2:30 PM and the algorithm finished overnight. 

- To be fair to both species, the same will be run for species P one time. Any runs after that need to be paired down for the small scope of this project. 

- Write cluster results to file:
```{r}
# inserting an 'S' into the csv file name
# write_csv(clusters_29, "../output/clusters_29_S.csv")
```

### Plot of Total Within Cluster Sum of Squares for Species S
```{r}
run_kmeans@metadata$tot_withinss %>% 
  as.data.frame() %>% 
  `colnames<-`("TOTWSS") %>% 
  mutate(K = 2:100) -> K_tot_withinss_S
```

```{r}
ggplot(K_tot_withinss_S, 
       aes(x = K, y = TOTWSS)) +
  geom_point() +
  ggtitle("Total Within Cluster Sum of Squares", 
          "Species S") +
  theme_bw() +
  xlab("K = Number of Clusters") + 
  ylab("Total Within Cluster SS") -> plot_totwss_s
```

```{r}
plot_totwss_s

```

## Inertia
```{r}
clusterInertia(profiles = tcounts(run_kmeans), # species S
               clusters = clusters(run_kmeans)
               ) -> inertia_S

length(inertia_S) # 29 clusters
```

### Plot of the Within Cluster Inertia for the "Best" Clustering Model of K = 29 for Species S
```{r}
# save cluster inertia data in df
tibble(
  "Cluster_Num" = 1:29, 
  "WC_Inertia" = inertia_S
) %>% 
  # better than base plot
  ggplot(aes(x = Cluster_Num,
             y = WC_Inertia)) +
  geom_point() +
  theme_bw() +
  ggtitle("Inertia Per Cluster", "'Best' Model, K = 29 Clusters for Species S") -> plot_k29_inertias_S

plot_k29_inertias_S


ggsave("../plots/plot_k29_inertias_S.png", 
       plot_k29_inertias_S, 
       device = "png", 
       width = 10, 
       height = 7)
```




## Saving Model Results in RDS file:
```{r}
# saveRDS(run_kmeans, 
#         file = "../output/run_kmeans_c29_S.rds")
```

- Testing Re-Import of Saved K-means Results from RDS File: 
```{r}
readRDS("../output/run_kmeans_c29_S.rds") -> test_import_RDS
```

```{r}
test_import_RDS
```

```{r}
test_import_RDS@metadata$tot_withinss %>% 
  as.data.frame() %>% 
  `colnames<-`("WSS")
```

## Additional Seeds Results
- Modifying the 'K' parameter to decrease run time
- Modifying the seed to pick different centers
```{r}
set.seed(23)
run_kmeans_2 <- coseq(s_bac_num,
                      K = 2:50,
                      transformation = "logclr",
                      model = "kmeans",
                      nstart = 100, 
                      iter.max = 1000)


```

```{r}
run_kmeans_2
```

```{r}
saveRDS(run_kmeans_2, "../output/run_kmeans_2.rds")
```

### Seed
- Modifying only the seed with same decreased K = 2:50 to reduce runtime but keep iterations high and nstart high. 
```{r, cache=TRUE}
set.seed(105)
run_kmeans_3 <- coseq(s_bac_num,
                      K = 2:50,
                      transformation = "logclr",
                      model = "kmeans",
                      nstart = 100, 
                      iter.max = 1000)
```


```{r}
```


```{r}
```

