---
title: "Coseq_P"
author: "Cassandra Sperow"
date: "2023-10-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(coseq)
```

## Experimental K-Means Clustering with Species P
- Using the R package 'coseq', perform K-means clustering on bacterial ASV data
```{r}
# read in the split dataset with only species P
read_csv("../output/p_bac.csv") -> p_bac
```

```{r}
dim(p_bac)
```


```{r}
head(p_bac)

p_bac[,-1] -> p_bac_num
```

## Run coseq() with K-means
- Uses MacQueen 1967 K-means algorithm 
- The below is commented out due to taking several hours to run. Results are saved in an RDS file in the output folder. 
- For additional seeds to compare, the cache = TRUE for code chunks that take a very long time to run. 
```{r}
# set.seed(42)
# run_kmeans_P <- coseq(p_bac_num, # numeric only
#                     K = 2:100,
#                     transformation = "logclr",
#                     model = "kmeans",
#                     nstart = 100,
#                     iter.max = 1000)
```

- To be able to knit without re-running the above several hours, the below is commented out.
```{r}
#run_kmeans_P
```
The above object returns: 
"An object of class coseqResults
 5889 features by 296 samples. 
 Models fit: K = 2 ... 100
 Chosen clustering model: K = 25"
 
 
 
## Save Model Results in RDS file:
```{r}
#saveRDS(run_kmeans_P, file = "../output/run_kmeans_P.rds")
```

## Re-import to be able to plot results:
```{r}
readRDS("../output/run_kmeans_P.rds") -> run_kmeans_P
```


```{r}
run_kmeans_P@metadata$tot_withinss %>% 
  as.data.frame() %>% 
  `colnames<-`("TOTWSS") %>% 
  mutate(K = 2:100) -> K_tot_withinss_P

K_tot_withinss_P
```

### Plot Total Within Sums of Squares and K Clusters
```{r}
ggplot(K_tot_withinss_P, 
       aes(x = K, y = TOTWSS)) +
  geom_point() +
  ggtitle("Total Within Cluster Sum of Squares", 
          "Species P") +
  theme_bw() +
  xlab("K = Number of Clusters") + 
  ylab("Total Within Cluster SS") -> plot_totwss_p
```


```{r}
plot_totwss_p
```

## Inertia
```{r}
clusterInertia(profiles = tcounts(run_kmeans_P), 
               clusters = clusters(run_kmeans_P)) -> inertia_P

length(inertia_P)
```

### Plot of the Within Cluster Inertia for the "Best" Clustering Model of K=25 for Species P
```{r}
# base plot
base::plot( 1:length(inertia_P), inertia_P, 
            main = "Within Cluster Inertia for K = 25")
```

## GGPlot Cluster Inertia for 'Best' K = 25s
```{r}
# better plot
tibble(
  "Cluster_Num" = 1:25, 
  "WC_Inertia" = inertia_P
) %>% 
  ggplot(aes(x = Cluster_Num, 
             y = WC_Inertia)) +
  geom_point() +
  theme_bw() +
  ggtitle("Inertia Per Cluster", "'Best' Model, K = 25 Clusters for Species P") -> plot_k25_inertias_P



plot_k25_inertias_P

ggsave("../plots/plot_k25_inertias_P.png", 
       plot_k25_inertias_P, 
       device = "png", 
       width = 10, 
       height = 7)

```


## Seed Experiments
- Setting cache = TRUE to be able to knit without re-running
- Modifying the 'K' parameter to decrease run time
- Modifying the seed to pick different centers with all other hyperparameters the same

### Seed 23
```{r, cache=TRUE}
set.seed(23)
run_kmeans_P2 <- coseq(p_bac_num, 
                       K= 2:50, 
                       transformation = "logclr",
                       model = "kmeans",
                       nstart = 100, 
                       iter.max = 1000)
```


```{r}
run_kmeans_P2
```


```{r}
saveRDS(run_kmeans_P2, "../output/run_kmeans_P2.rds")
```


### Seed 105
- Modifying only the seed with same decreased K = 2:50 to reduce runtime but keep iterations high and nstart high. 
```{r, cache=TRUE}
set.seed(105)
run_kmeans_P3 <- coseq(p_bac_num, 
                       K= 2:50, 
                       transformation = "logclr",
                       model = "kmeans",
                       nstart = 100, 
                       iter.max = 1000)

```


```{r}
run_kmeans_P3
```


```{r}
saveRDS(run_kmeans_P3, "../output/run_kmeans_P3.rds")
```

### Seed 12
```{r, cache=TRUE}
set.seed(12)
run_kmeans_P4 <- coseq(p_bac_num, 
                       K = 2:50, 
                       transformation = "logclr", 
                       model = "kmeans", 
                       nstart = 100,
                       iter.max = 1000)

```



```{r}
run_kmeans_P4
```


```{r}
saveRDS(run_kmeans_P4, "../output/run_kmeans_P4.rds")
```



### Compare Results from 3 Seed Experiments for Species P




