---
title: "Coseq"
author: "Cassandra Sperow"
date: "2023-10-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

### Coseq is an R library that works with count-based data to implement clustering with better performance on compositional data
- Coseq offers choices of data transformations depending on one's use-case
- The ASV counts are by nature compositional data
- Compositional data are linearly dependent, violating many assumptions from traditional statistical methods. 
- See publication: https://arxiv.org/pdf/1704.06150.pdf

#### Read in count-based data and transpose according to what the package documentation states: 
- Takes a matrix or data.frame of gene-level or ASV-level counts of dimension n x d 
- where n = genes or ASVs as rows and d = samples as columns: 
```{r}
read_csv("../output/rev_bacteria.csv") -> rev_bacteria

rev_bacteria %>% 
  t() %>% 
  as.data.frame() %>% 
  janitor::row_to_names(1) %>% 
  map(., ~as.numeric(.)) %>% 
  as.data.frame() -> t_rev_bacteria_counts
```

- Inspect transposed data frame
- all numeric
- raw counts
- n = ASVs in rows, i.e., the bacteria are the observations
- d = samples in columns
- See documentation for why it needs to be in this format: 
https://www.bioconductor.org/packages/devel/bioc/vignettes/coseq/inst/doc/coseq.html
```{r}
# no ASV name column
head(t_rev_bacteria_counts)
```
## Write transposed raw counts df to file for usage later
- First, make sure to have the ASV names:
```{r}
# but first get names of ASVs into column
names(rev_bacteria[-1]) -> names_column

bind_cols(ASV = names_column, 
          t_rev_bacteria_counts) -> t_rev_bacteria_counts

write_csv(t_rev_bacteria_counts, "../output/t_rev_bacteria_counts.csv") # no ASV name column but they are in original order

```

# Quick Start Guide for ```coseq``` package
- It may necessary for new users to go through the commands on this page to set up Bioconductor and coseq:
https://www.bioconductor.org/install/
- Number of K and max iterations are low to try out method first
Note:
## Optional start here, read in in previously transposed data
```{r}
# includes ASV name column
readr::read_csv( "../output/t_rev_bacteria_counts.csv") -> t_rev_bacteria_counts  

```

```{r}
head(t_rev_bacteria_counts)
```

## LogCLR with K-Means using ```coseq```
- Log centered log ratio (logCLR) is recommended by the researchers and writers of the coseq package for compositional data

```{r}
library(coseq)
# try the default method suggested by documentation
# number of iter and number of clusters is low to try out method
coseq(t_rev_bacteria_counts[, -1], # take out ASV names in 1st column
      K=2:25, # experiment with low number
      transformation="logclr", # recommended from research article
      norm="TMM", 
      meanFilterCutoff=50, 
      model="kmeans",
      nstart=1, 
      iter.max=10) -> coseq_out
```


```{r}
# summary
summary(coseq_out)
```


```{r}
clusters(coseq_out) 
```

- There are two species of coral in dataset
- Sometimes the algorithm finds 2 clusters, depending on the random start. 
- Re-running without adjusting paramters and using 'nstart' with higher number will always give randomly different results; however setting the seed is a bit artificial until we can get somewhat more stable results. 


```{r}
# conditional probabilities of cluster membership
assay(coseq_out) %>% as.data.frame()
```


- Per article below, for K-Means, "...the use of untransformed or logCLR-transformed data with K-means thus appears to be a more coherent choice." pg. 12
link: https://arxiv.org/pdf/1704.06150.pdf

### CLR Transformation
- Dr. C advised to use this transformation from the raw counts to the centered log ratio (CLR)
- Note: Increasing the max iterations
```{r}
# changed transformation type
coseq(t_rev_bacteria_counts[,-1],
      K=2:25, 
      transformation="clr", # centered log ratio
      norm="TMM", # bio-specific setting for these data
      meanFilterCutoff=50, 
      model="kmeans",
      nstart=1, 
      iter.max=10) -> coseq_out_clr
```
 
```{r}
summary(coseq_out_clr)
```


```{r}
# cluster results
clusters(coseq_out_clr) 
```

```{r}
# conditional probabilities of cluster membership
assay(coseq_out_clr) %>% as.data.frame()
```


-- Changing max iterations to 50 as an experiemnt
```{r}
coseq(t_rev_bacteria_counts[,-1],
      K=2:25, 
      transformation="clr", # centered log ratio
      norm="TMM", # bio-specific setting for these data
      meanFilterCutoff=50, 
      model="kmeans",
      nstart=1, 
      iter.max=50) -> coseq_out_clr_50

summary(coseq_out_clr_50)
```


- Discussion and Notes with Client on 10-16-2023: 
- Mention that coseq() uses MacQueen K-means, 1967. 
- Mention for further research that different K-means can give different results and can be researched and optimized further.
- Since the question is to find more specific groups of bacteria, and since the two species in the dataset may tend to give only 2 clusters, split the data into 2 datasets, one for each species. 
- To show the variability of the results, use about 5 random seeds per species for anything to be reproducible and to compare. 
- Please see the additional .Rmd files for the split and further analysis. 
