---
title: "Data_Split_Species"
author: "Cassandra Sperow"
date: "2023-10-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Split Transposed Dataset into Two Datasets: One for Each Species

```{r}
# read in transposed counts df
read_csv("../output/t_rev_bacteria_counts.csv") -> t_rev_bacteria_counts
```

### Select the ASV names column with only the columns that begin with 'S' for the species ......
- Save to object
- Note that there are 363 columns for species S
```{r}
# sample IDs that begin with 'S' are the species .....
t_rev_bacteria_counts %>% 
  select(1, starts_with("S")) -> s_bac
```

### Select the ASV names column with only the columns that begin with 'P' for the species ......
- Save to object 
- Note there are 296 columns for species P
```{r}
t_rev_bacteria_counts %>% 
  select(1, starts_with("P")) -> p_bac
```

## Data Check Species S
```{r}
str(s_bac)
```



## Data Check Species P

```{r}
str(p_bac)
```



## Write each to file

```{r}
write_csv(s_bac, "../output/s_bac.csv")

write_csv(p_bac, "../output/p_bac.csv")
```


