---
title: "Modeling, Clustering, and Final Analysis"
author: "Cassandra Sperow"
date: "2023-11-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(library(tidyverse))
```

## Revised Final Analysis Combining Ridge 200 Modeling Predictors, Cluster Assignments & Bacteria Taxonomy

```{r}
# read in modeling results from Final Ridge & Lasso Modeling Revised Rmd file
readRDS("../output/ridge_coefs.rds") -> ridge_coefs
# read in cluster assignments from K-means in 'coseq' files
read_csv("../output/species_P_cluster_assignments.csv") -> clusters_P
read_csv("../output/species_S_cluster_assignments.csv") -> clusters_S
# read in bacteria taxonomy table
read_delim("../data/Files/16S_tax.txt") -> bac_tax

```

## Join the coefficients with the cluster assignments for each species and combine
```{r, joins-ridge-200}
################### combine ridge 200 predictors with cluster assignments and taxonomy
# species P
ridge_coefs[3] %>% # Ridge 200
  as.data.frame() %>% 
  filter(str_detect(predictor, "ASV")) %>% 
  left_join(clusters_P, join_by("predictor" == "ASV")) %>% 
  left_join(bac_tax, join_by("predictor" == "ASV")) %>% 
  group_by(Family, Genus) %>%  # 16 groups
  ungroup() %>% 
  mutate(coral_species = "P") -> predictors_p_df

# species S
ridge_coefs[3]  %>% 
  as.data.frame() %>% 
  filter(str_detect(predictor, "ASV")) %>% 
  left_join(clusters_S, join_by("predictor" == "ASV")) %>% 
  left_join(bac_tax, join_by("predictor" == "ASV")) %>% 
  group_by(Family, Genus) %>%  # 16 groups
  ungroup() %>% 
  mutate(coral_species = "S") -> predictors_s_df

# combine
bind_rows(
predictors_p_df, 
predictors_s_df) -> predictors_clade_c_ridge # genus or family are best per client.

write_csv(predictors_clade_c_ridge, "../output/predictors_clade_c_ridge.csv")
```


```{r}
predictors_clade_c_ridge %>% # 400 because the k-means was done for each species separately
  arrange(-abs(s1)) %>% # in order of coefficient absolute value: high to low
  # since cluster assignments were by species, the join added double columns for the second coral species
  # filtering here gives the unique list
 # filter(coral_species == "P") %>% # 
  rename(coef = s1) %>% 
 # filter(coef > 0) %>% 
  #filter(predictor %in% c("ASV1321", "ASV3327" ))
 # filter(seed12_run4 == 1 | seed12_run4==6 ) %>%  # or for species S:   seed12_run4==6
  select(predictor, Family, seed12_run4)
```

## What is the Order or Family that's most present?
```{r}

predictors_clade_c_ridge %>% # 400 because the k-means was done for each species separately
  arrange(-abs(s1)) %>% 
  filter(coral_species == "P") %>% 
  rename(coef = s1) %>% 
  group_by(Family) %>% 
  summarise(n = n()) %>% 
  arrange(-n, Family) %>% 
  mutate(proportion = n/sum(n))
```
- About 19 % are of the family Alteromonadaceae, the most present family in the ridge coeficients list. Since the 200 ASVs used for this model were from PCA results

## Plots with varying taxonomies
```{r, plots-ridge-200}

# predictor cluster assignments from ridge 200
# facet wrap for coral species
predictors_clade_c_ridge %>% 
  ggplot(aes(x = seed12_run4, fill = Family)) +
  geom_histogram(stat = 'count', show.legend = FALSE) +
  viridis::scale_fill_viridis(discrete = T) +
  theme_bw() +
  xlab("Cluster Assignment in Seed 12 K-Means") + 
  ggtitle("Cluster Assignnments of Ridge 200 Predictors with Taxonomy: Family") +
  facet_wrap(~paste("Species ", coral_species)) 
# facet wrap for coral species
predictors_clade_c_ridge %>% 
  ggplot(aes(x = seed12_run4, fill = Genus)) +
  geom_histogram(stat = 'count', show.legend = FALSE) +
  viridis::scale_fill_viridis(discrete = T) +
  theme_bw() +
  xlab("Cluster Assignment in Seed 12 K-Means") + 
  ggtitle("Cluster Assignnments of Ridge 200 Predictors with Taxonomy: Genus") +
  facet_wrap(~paste("Species ", coral_species)) 



################### NO facet wrap - BOTH species together
predictors_clade_c_ridge %>% 
  ggplot(aes(x = seed12_run4, fill = Family)) +
  geom_histogram(stat = 'count', 
                 #show.legend = FALSE
                 ) +
  viridis::scale_fill_viridis(discrete = T) +
  theme_bw() +
  xlab("Cluster Assignment in Seed 12 K-Means") + 
  ggtitle("Cluster Assignnments of Ridge 200 Predictors with Taxonomy: Family") +
  theme(legend.key=element_blank(), legend.key.size=unit(5,"point")) -> plot_ridge_200_family
  #guides(colour=guide_legend(nrow=10)) 
  

ggsave(filename = "../plots/plot_ridge_200_family.png",
       plot = plot_ridge_200_family, 
       device = "png",
       width = 17,
       height = 11,
       units = "in",
       )

################### facet wrap for species
predictors_clade_c_ridge %>% 
  ggplot(aes(x = seed12_run4, fill = Family)) +
  geom_histogram(stat = 'count', 
                 show.legend = FALSE
                 ) +
  viridis::scale_fill_viridis(discrete = T) +
  theme_bw() +
  xlab("Cluster Assignment in Seed 12 K-Means") + 
  ggtitle("Cluster Assignnments of Ridge 200 Predictors with Taxonomy: Family") +
  theme(legend.key=element_blank(), legend.key.size=unit(5,"point")) +
  facet_wrap(~paste("Species ", coral_species)) 


# 
# predictors_clade_c_ridge %>% 
#   ggplot(aes(x = seed23_run2, fill = Genus)) +
#   geom_histogram(stat = 'count', show.legend = FALSE) +
#   viridis::scale_fill_viridis(discrete = T) +
#   theme_bw() +
#   xlab("Cluster Assignment in Seed 23 K-Means") + 
#   ggtitle("Cluster Assignnments of Ridge 200 Predictors with Taxonomy: Genus") 

###################  facet wrap 
predictors_clade_c_ridge %>% 
  ggplot(aes(x = seed12_run4, fill = Order)) +
  geom_histogram(stat = 'count', 
                 #show.legend = FALSE
                 ) +
  viridis::scale_fill_viridis(discrete = T) +
  theme_bw() +
  xlab("Cluster Assignment in Seed 12 K-Means") + 
  ggtitle("Cluster Assignnments of Ridge 200 Predictors with Taxonomy: Order") + 
  theme(legend.key=element_blank(), legend.key.size=unit(5,"point")) +
  guides(colour=guide_legend(nrow=30)) +
  facet_wrap(~paste("Species ", coral_species)) 
```


# ROC for recommended model
```{r}
ridge_coefs[3]
```

