---
title: "Algae (ITS2) Table Exploratory Data Analysis"
author: "Cassandra Sperow"
date: "2023-09-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(library(tidyverse))
```
# Questions:

##### How many algal ITS2 types?  89

##### How many coral samples? 659 

##### What are distributions like per category or per variable?


## Algae (ITS2 Table)
- Are the coral sample IDs in rows? Yes
- What is format of dataset? 659 by 89 where 88 columns are ITS2 IDs

```{r}
# separator is \t
# no file extension
read_delim("../data/Files/ITS2_table") %>% 
   # Notice the coral sample ID has the Red Sea region acronym, reef number 
  rename(coral_sample_id = `...1`) -> algae

dim(algae) # 659  89

algae
```

- Are the columns all unique? Yes, they are each an ITS2 sequence type
- There are 89 algal ITS2 sequence types
```{r}
unique(names(algae))
```

### Algae Taxonomy
#### Clade Observations
```{r}
read_delim("../data/Files/ITS2_tax.txt") -> al_tax

al_tax # 88 by 5
```

- How many clades are there? about 5
- Dr. C said that it would be good to do analysis based on clade level. 

#### Unique clades: qty 5: **A, B, C, D, G**
```{r}

al_tax %>% 
  group_by(Clade) %>% 
  summarise(n = n()) ### samples per clade are not as evenly stratified
```

- How many algae species? 
- There are 5 different species listed with some algae having a ```None``` value 
- These were a part of the findings from the published article where some might not have been ID'd with a specific species name when their genetic profile was analyzed
```{r}
al_tax %>% 
  group_by(`Associated species`) %>% 
  summarise(n = n())
```
- How many majority sequences are there in the taxonomy: 32 different majority sequences
```{r}
al_tax %>% 
  group_by(`Majority ITS2 sequence`) %>% 
  summarise(n = n())
```

# Joining Algae ITS2 Counts per sample ID with clade, majority sequence, and sequence profile columns from the taxonomy table to have all information per coral sample ID: 

- The only common column is ITS2 type ID
- Start with algae ITS2 table:  Take table with coral sample IDs and pivot to where they can be joined based on ITS2 (there is only one clade, majority, and profile per ITS2 and each coral sample has all ITS2s, of which there are 88)

```{r}


algae %>% 
  # coral sample IDs are in 1 column
  # pivot to where the coral samples have all of the ITS2s 
  # it makes a lot of rows, but need this step in order to pivot again
  pivot_longer(cols = starts_with("X")) %>% 
  # with ITS2s now in their own column, need to pivot again to get samples as headers
  pivot_wider(names_from = coral_sample_id, 
              values_from = value) -> algae_pv

# The ITS2s are pivoted into one column but it still have an "X" in them
head(algae_pv)
```

- Take out "X" in order to join by ITS2 and get clade per sample
```{r}
# Before joining, need to make ITS2 columns the same data type in the taxonomy table
al_tax %>% 
  mutate(ITS2 = as.character(`ITS2 type profile UID`), 
         .before = `ITS2 type profile UID`) -> al_tax


algae_pv %>% 
  # separating out the "X" from the ITS2
  separate(name, into = c("a", "ITS2"), sep = "X") %>% 
  #select(`PKAU-R3-12`) # yes checks out
  select(-a) %>% 
  inner_join(al_tax, by = c("ITS2" = "ITS2")) -> algae_coral_clade_maj


# testing some of the columns to make sure they joined with correct sample ID
# picking 3 in the middle of one of the original files from Dr. C to test
# yes, these match up with picking them randomly and checking original tables. 
# algae_coral_clade_maj %>% 
# select(`PKAU-R3-12`, `PKAU-R1-20`, `PWAJ-R2-14`, ITS2, Clade)


```



#### Before moving on, exploring quick plots below to see the data by Clade

```{r}
algae_coral_clade_maj %>% 
  ggplot(aes(x = Clade, color = Clade, fill = Clade)) +
  geom_bar() +
  theme_bw() + 
  ggtitle("Bar Chart of Clade Counts")
```

```{r}
algae_coral_clade_maj %>% 
  ggplot(aes(x = `Majority ITS2 sequence`, color = Clade, fill = Clade)) +
  geom_histogram(stat="count") +
  theme_bw() +
  ggtitle("Count of Majority ITS2 Sequence with Clade") +
  coord_flip()

```

- Need to pivot back with samples as rows
```{r}
algae_coral_clade_maj %>% 
  # take out associated psecies column for now
  select(-`Associated species`) %>% 
  # there are two ITS2 columns each of a diff data type
  # only need the one that is numeric
  select(-ITS2) %>% 
  # simplify the pivot by taking out the longer profile (analysis first on clade)
  select(-`ITS2 type profile`) %>% 
  # simply names
  rename(ITS2 = `ITS2 type profile UID`, 
         Majority = `Majority ITS2 sequence`) %>% 
  # check names
 # names()
  # pivot back to have samples as rows
  pivot_longer(cols = 1:659) %>% 
  # rename 'name' and 'value' to be clearer
  rename(sample_id = name, 
         its2_count = value) %>% 
  # arrange by sample name
  arrange(sample_id) -> algae_df

## Quality Checks: 
  # # group by sample name to check info per sample
  # group_by(name) %>% 
  # summarize(n = n()) #%>% # yes, there are 88 ITS2 types and each sample has 88


glimpse(algae_df)
  
```

## Compare new data frame with original for data quality check
- Randomly picking 3 sample names to check with original algae table
```{r}
algae_df %>% 
 # filter(sample_id=="PDOG-R1-1") # yes, this is right
  #filter(sample_id=="PWAJ-R3-10") # yes this is right
  filter(sample_id=="PWAJ-R1-6") # yes, this is right
```

## More quality checks before creating new columns for region, reef, species
Yes, there are 659 groups of ```sample_id```
Yes, there are 88 unique ITS2 codes
```{r}
algae_df %>% 
  group_by(sample_id) # yes 659 unique

algae_df %>% 
  group_by(ITS2) # yes 88 unique
```

#### Create columns with other categorical data of region, reef, and species
- This is easier than joining with metadata because sample names already have species, region, and reef within acronyms
- Save as ```algae_df```
```{r}
algae_df %>% 
  # extract letter P or S at beginning of sample name for which species
  mutate(species = str_sub(sample_id, 1, 1), 
         region = str_sub(sample_id, 2,4),
         reef = str_sub(sample_id, 6,7)) %>% 
  # ITS2 id should be a categorical not a double 
  mutate(ITS2 = as.character(ITS2))-> algae_df
```

```{r}
glimpse(algae_df)
```


# Plots

## Quick snapshot of main variables with colors by Clade
```{r}
GGally::ggpairs(select(algae_df, -1, -3, -4, -8), mapping = aes(color = Clade)) -> ggpairs_color_clade

ggpairs_color_clade
```

- Saving plot
```{r}

ggsave("GGpairs_color_clade.png", ggpairs_color_clade, device = "png", 
       path = "../plots")
```

## Quick snapshot of main variables with colors by Species

```{r}
GGally::ggpairs(select(algae_df, -1, -3, -4, -8), mapping = aes(color = species)) -> ggpairs_color_species

ggpairs_color_species
```

- Saving plot
```{r}
ggsave("GGpairs_color_species.png", ggpairs_color_species, device = "png", path = "../plots")
```


```{r}
algae_df %>% 
  arrange(-its2_count) %>% 
  ggplot(aes(x = its2_count, y = Clade, color = Majority )) +
  geom_point() + 
  theme_bw() +
  ggtitle("ITS2 Counts by Clade & Majority Sequence")
```
- From the above, there are many different profiles that fall under Clades A, C and D with sometimes only 1 or a few different majorities for B, D and G. 


### Scatter Plot(s)
```{r}
#plot(algae_df$its2_count)

ggplot(algae_df, aes(x = 1:nrow(algae_df), y = its2_count, color = Clade)) +  # use nrow function to have the index values on the x axis 
  geom_point() +
  theme_bw() +
  xlab("Index") +
  ylab("ITS2 Count (Abundance)") +
  ggtitle("Scatter Plot of ITS2 Counts by Clade") -> scatter_its2

scatter_its2

ggsave("scatter_its2.png", scatter_its2, device = "png", path = "../plots")
  
```
- What if I were to log the count variable?
```{r}
ggplot(algae_df, aes(x = 1:nrow(algae_df), y = log(its2_count), color = Clade)) +  # use nrow function to have the index values on the x axis 
  geom_point() +
  theme_bw() +
  xlab("Index") +
  ylab("Log of ITS2 Count (Abundance)") +
  ggtitle("Scatter Plot of log(ITS2 Count) by Clade")
```


### What does data frame look like if I filter out ITS2 counts that are 0?
- Compare with similar plot
```{r}
algae_df %>% 
  filter(its2_count >0) %>% # 778 rows x 8 
  ggplot(aes(x = 1:nrow(.), y = its2_count, color = Clade)) +  # use nrow function to have the index values on the x axis 
  geom_point() +
  theme_bw() +
  xlab("Index") +
  ylab("ITS2 Count (Abundance)") +
  ggtitle("Scatter Plot of ITS2 Counts > 0") -> scatter_its2_gr_0

scatter_its2_gr_0

ggsave("scatter_its2_gr_0.png", scatter_its2_gr_0, device = "png", path = "../plots")

```
- From the above, it may be useful to test with Clade A first

### Plots for Majority Sequence

```{r}
algae_df %>% 
  group_by(Majority) %>% # 32 groups
  ggplot(aes(y = Majority, color = Clade, fill = Clade)) +
  geom_bar() +
  facet_wrap(~species) +
  theme_bw() +
  ggtitle("Majority Sequence by Species & Clade", "P = P. verrucosa, S = S. pistillata") -> plot_ct_maj_clade_species


plot_ct_maj_clade_species

ggsave("Plot_ct_maj_clade_species.png", device = "png", path = "../plots")
```
### Number of rows per species per clade
- How are observations (rows) split between species and clade if Clade A is used for testing? Or if minor clades are filtered out?
```{r}

algae_df %>% 
  group_by(species, Clade) %>% 
  summarise(n = n()) %>% 
  arrange(-n)
```

- When grouping by sample ID and Clade, many samples have 50 rows of Clade A, 32 or so of Clade C, and only 1-4 of the other Clades. 
- Within the 50, one sample id has several majority types with different ITS2 types but only one with have any value of count (abundance)
```{r}
algae_df %>% 
  # group by sample id and clade to see how many rows per sample per clade
  group_by(sample_id, Clade) %>% 
  # put in order of sample id, then its2 count to see larges to smallest counts per sample id
  arrange( sample_id, -its2_count)
  # summarise(n = n()) %>%
  # arrange(sample_id)
```

- Where are highest ITS2 counts?
- Should ITS2 counts greater than 0 be taken out? 
```{r}
algae_df %>% 
  arrange(-its2_count) %>% 
  filter(its2_count>0) # 778 rows left
 
```


# Write csv file 
- Next step is to join with bacteria data, pending questions of perhaps parsing down clades and/or ASVs depending on getting more familiar with ASV table
```{r}
write_csv(algae_df, "../output/algae_df.csv")
```



# Filtering out ITS2 counts that are 0
- And regraphing to get a quick snapshot since distributions where skewed with so many 0 counts
```{r}
algae_df %>% 
  filter(its2_count >0) %>% 
  # checking to see how many samples
  group_by(sample_id) %>% # 658 down 1 from 659
  ungroup() -> algae_df_no_
  # redoing plot of the 'quick snapshot' with ggpairs to see distributions
  GGally::ggpairs(columns = c(2, 5, 6,7), 
                  mapping = aes(color = species))
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```

