---
title: "Jacard Index for K-Means Clustering Consistency"
author: "Cassandra Sperow"
date: "2023-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Testing Jacard Index 

```{r}
# Sample cluster assignments for three runs
set.seed(123)  # For reproducibility
cluster_run1 <- kmeans(mtcars, centers = 3, nstart = 1)$cluster
cluster_run2 <- kmeans(mtcars, centers = 3, nstart = 1)$cluster
cluster_run3 <- kmeans(mtcars, centers = 3, nstart = 1)$cluster

```


```{r}
# Function to calculate Jaccard index
jaccard_index <- function(cluster1, cluster2) {
  intersection <- length(intersect(cluster1, cluster2))
  union <- length(union(cluster1, cluster2))
  return(intersection / union)
}

# Calculate Jaccard index for each pair of runs
jaccard_12 <- jaccard_index(cluster_run1, cluster_run2)
jaccard_13 <- jaccard_index(cluster_run1, cluster_run3)
jaccard_23 <- jaccard_index(cluster_run2, cluster_run3)

cat("Jaccard Index (Run 1 vs Run 2):", jaccard_12, "\n")
cat("Jaccard Index (Run 1 vs Run 3):", jaccard_13, "\n")
cat("Jaccard Index (Run 2 vs Run 3):", jaccard_23, "\n")

```


# Jacard Index for Bacteria ASV K-Means Clustering Seed Experiments
- See also Rmd files with 'Coseq' in title for the individual runs of K-means per species. 

## Read in Cluster Assignments
- Cluster assignments were found in the 'coseq' files for each species. 
- 3 random runs per species were used to see if the clustering assignments and groups would be the same, and if so, how much the same or different? 
```{r}
# import the cluster assignments that were saved from the previous 'coseq' files
read_csv("../output/species_S_cluster_assignments.csv") -> clusters_S
read_csv("../output/species_P_cluster_assignments.csv") -> clusters_P

```

# Species S 
```{r}
# save into simple variables
clusters_S$seed23_run2 -> s1
clusters_S$seed105_run3 -> s2
clusters_S$seed12_run4 -> s3
```


```{r}
# Function to calculate Jaccard index
jaccard_index <- function(cluster1, cluster2) {
  intersection <- length(intersect(cluster1, cluster2))
  union <- length(union(cluster1, cluster2))
  return(intersection / union)
}

# Calculate Jaccard index for each pair of runs
jaccard_12 <- jaccard_index(s1, s2)
jaccard_13 <- jaccard_index(s1, s3)
jaccard_23 <- jaccard_index(s2, s3)

cat("Jaccard Index (Run 1 vs Run 2):", jaccard_12, "\n")
cat("Jaccard Index (Run 1 vs Run 3):", jaccard_13, "\n")
cat("Jaccard Index (Run 2 vs Run 3):", jaccard_23, "\n")
```

- The Jacard calcuations indicate that the cluster assignments for runs 2 and 3 are more similar than the other run. This makes sense as there were two runs that had a similar number of K clusters, which were runs 2 and 3 where the K clusters were 8 and 7, respectively. The other run found K=13 clusters, which is indicated by the Jacard Index value of 0.471 and 0.412, respectively, for 1 vs 2 and 1 vs. 3. 

# Species P
```{r}
clusters_P$seed23_run2 -> p1
clusters_P$seed105_run3 -> p2
clusters_P$seed12_run4 -> p3
```


```{r}
# Function to calculate Jaccard index
jaccard_index <- function(cluster1, cluster2) {
  intersection <- length(intersect(cluster1, cluster2))
  union <- length(union(cluster1, cluster2))
  return(intersection / union)
}

# Calculate Jaccard index for each pair of runs
jaccard_12 <- jaccard_index(p1, p2)
jaccard_13 <- jaccard_index(p1, p3)
jaccard_23 <- jaccard_index(p2, p3)

cat("Jaccard Index (Run 1 vs Run 2):", jaccard_12, "\n")
cat("Jaccard Index (Run 1 vs Run 3):", jaccard_13, "\n")
cat("Jaccard Index (Run 2 vs Run 3):", jaccard_23, "\n")
```

- Species P resulted in very consistent clustering indicated by the Jaccard Index values above. 
- Species S has less consistent clustering results also indicated by the Jaccard Index values above for Species S. 

```{r}
```

