---
title: "Join"
author: "Cassandra Sperow"
date: "2023-10-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = T)
library(tidyverse)
```

# Join Bacteria Table with Algal Clade

**Revised 10-30-2023 with section for prioritizing clade C. Please scroll down to section titled 'Clade C Prioritization' **

- How can we explain which bacteria are associated with each clade? 

- Can we learn more about the relationship between algal clade based on bacteria present in the given dataset of samples?

- In order to address these questions, joining the clode per sample ID with the bacteria ASV table is necessary.

### Read in Bacteria Table (ASVs)

- Start with reduced ASV table of original counts-based data. This table was filtered such that any given ASV is found in at least 3 coral samples: 
```{r}
read_csv("../output/rev_bacteria.csv") -> rev_bacteria
```

- Checking data
```{r}
head(rev_bacteria)
```


### Read in Algae Table (ITS2s)

- The algae_df.csv file was made from the original ITS2 table and includes information from the metadata of region, majority sequence, majority sequence type, clade, and other variables:

```{r}
read_csv("../output/algae_df.csv") -> algae_df
```

- Checking data
```{r}
dim(algae_df)
head(algae_df)
```

- Examine the algal counts for two samples: 
```{r}
algae_df %>% 
  filter(sample_id == "PDOG-R1-1") %>% 
  arrange(-its2_count) %>% 
  group_by(Clade)


algae_df %>% 
  filter(sample_id == "SWAJ-R3-21") %>% 
  arrange(-its2_count, Clade) %>% 
  group_by(Clade)
```
- From the above, sample PDOG-R1-1 has the majority clade of A where the ITS2 abundance count is 21,483, and the rest of the ITS2 counts are 0. 

- For sample SWAJ-R3-21, the max ITS2 count is 20894, which is clade A. There are ITS2 counts for clade B, C, D. Some samples have ITS2 counts for more than one clade.


**Revision from client: Focus on clade C because the coral researching community knows less about clade C than A.** 

- Need to obtain the sampling data for clade C to see how many observations there are for clade C before modeling. 

- Group by sample id and clade. 

- For the scope of this project, focus on modeling clade C. For ex, sample SYAN-R4-22 has ITS2 counts for clade A and C whereas most samples have only one clade. For this sample, take the clade C information. 

- After seeing how many observations have clade C sequencing data, can balance with observations from other clades for binary classifciation modeling. 
 
```{r}
# reviewing groups of sample ID and Clade
algae_df %>% 
  group_by(sample_id, Clade)
```
 - From the above, there are many, many rows with 0 values for ITS2 counts because not every algae type was found in every sample.

```{r}
algae_df %>% 
  group_by(sample_id, Clade) %>% # yes, 659 samples
  # create new column of the majority clade's its2 counts (the max since rest are 0)
  # this column is no longer needed since lcient advised to focus on clade C ---Please see revised data wrangling for clade C below
  mutate(max_clade_count = max(its2_count)) %>% 
  # ungroup data frame
  ungroup() %>% 
  # filter out the 0 values per sample
  filter(its2_count != 0) %>% # 778 rows, so some samples are giving extra rows
  filter(sample_id == "SYAN-R4-22")
```

- Filtering out the secondary clades as mentioned above: 

**Revision 10-30-2023: Client advises to focus on clade C; therefore, please see below section on wrangling data to prioritize clade C not whatever the dominant clade is.**

```{r}
algae_df %>% 
  group_by(sample_id, Clade) %>% # yes, 659 samples
  # create new column of the majority clade's its2 counts (the max since rest are 0)
  mutate(max_clade_count = max(its2_count)) %>% 
  # ungroup data frame
  ungroup() %>%  # 57,992  rows
  # filter out where the its2 count is 0
  filter(its2_count > 0) %>%  # 778 rows - therefore, some samples have two clades
  # arrange by the max_clade_count in order to slice per group
  arrange(-max_clade_count) %>% 
  # in order to slice the max clade count per group, need to group by sample id
  group_by(sample_id) %>%  # 658 samples, so somewhere, one of them has a 0 max count
  # based on arrangement above, this will grab max clade count row per sample if more than one. 
  dplyr::slice(1) %>% 
  # don't need this columns any longer as it's now the same as its2 count
  select(-max_clade_count) %>% 
  # reorder 
  select(sample_id, Clade, 
         ITS2, its2_count, 
         Majority, species, 
         region, reef) -> algae_clade_df


algae_clade_df %>% 
  # check that it sliced the right row for sample SYAN-R4-22
  filter(sample_id == "SYAN-R4-22")
  
```

```{r}
dim(algae_clade_df)
head(algae_clade_df)
```

- Write this to csv for later if needed
```{r}
write_csv(algae_clade_df, "../output/algae_clade_df.csv")
```

### Join Based on Sample ID

- Since there is one less sample in the algae df, use this with a left join to match up ASV columns for the 658 samples.
- Note: The ASV table was previously filtered down from ~33K to ~5,889 ASVs based on filtering criteria from Dr. Cárdenas. 

```{r}
algae_clade_df %>% 
  left_join(rev_bacteria, by = join_by(sample_id)) -> coral_data # joined algae and bacteria data
```

- Examine joined data
- Note that ITS2 should be a category not a numeric
```{r}
coral_data %>% 
  mutate(ITS2_f = as.factor(ITS2), .after = its2_count) %>% 
  select(-ITS2) -> coral_data

head(coral_data)
```

- Write to file
```{r}
write_csv(coral_data, "../output/coral_data.csv")
```


# Clade C Prioritization

As of 10-30-2023, client advised to focus on clade C (not clade A which was the dominant clade). 

The below data wrangling obtains the clade C information per sample if the sample had any sequencing data for clade C. 

Considerations: 

- Filter for Clade C info and examine how many observations there are, especially when splitting by species because each species has a different genetic profile that was previously obtained in: 

Buitrago-López, Cárdenas, et al. (2023). Disparate population and holobiont structure of pocilloporid corals across the Red Sea gradient demonstrate species-specific evolutionary trajectories. Molecular Ecology. 2023;32:2151–2173. DOI: 10.1111/mec.16871 

- Species matters. The statistical and machine learning modeling will be later split by species because they each have different characteristics. 

## Filtering for Clade C and taking out 0 values for ITS2 count 

#### Clade C has 139 rows with non-zero values. This is extremely small for machine learning. This will be balanced with non-C clade observations 
```{r}
# use algae data frame wrangled from 'ITS2 Table.Rmd'
algae_df %>% 
  # prioritize clade C observations for client
  filter(Clade == "C") %>% 
  # remove 0 counts
  filter(its2_count != 0) %>%  # 139 rows 
  # group to see how many sample ID have clade C sequencing
  group_by(sample_id) # 111 sample IDs have sequencing data for clade C
```

- Some samples have more than one clade C sequence. Ex: PDOG-R2-22 has genetic sequencing for C3z and C1S.  These will be treated as separate observations because each alphanumeric sequence is a different algae type within the wider clade category. 

## Redo Join 
```{r}
algae_df %>% 
  # remove 0 counts
  filter(its2_count != 0) %>% 
  left_join(rev_bacteria, by = join_by(sample_id)) %>% 
  # clarify that ITS2 should be a type for anyone reading it in later
  rename(ITS2_type = ITS2) -> rev_coral_join
```

## Check a few samples with original tables

"PYAN-R3-1" - Verified
"PDOG-R1-7" - Verified
"SFAR-R1-14" - Verified

```{r}
rev_coral_join %>% 
 # filter(sample_id == "PYAN-R3-1") %>% # yes both algae and bacteria counts joined correctly from original tables
 # filter(sample_id == "PDOG-R1-7") # yes, this one also checks out with original tables
  filter(sample_id == "SFAR-R1-14") # yes

```

## Write new data frame to csv for other analyses 
```{r}
write_csv(rev_coral_join, "../output/rev_coral_join.csv")
```






