---
title: "Join"
author: "Cassandra Sperow"
date: "2023-10-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Join Bacteria Table with Algal Clade

- How can we explain which bacteria are associated with each clade? 

- Can we learn more about the relationship between algal clade based on bacteria present in the given dataset of samples?

- In order to address these questions, joining the clode per sample ID with the bacteria ASV table is necessary.

### Read in Bacteria Table (ASVs)

- Start with reduced ASV table of original counts-based data. This table was filtered such that any given ASV is found in at least 3 coral samples: 
```{r}
read_csv("../output/rev_bacteria.csv") -> rev_bacteria
```

- Checking data
```{r}
head(rev_bacteria)
```


### Read in Algae Table (ITS2s)

- The algae_df.csv file was made from the original ITS2 table and includes information from the metadata of region, majority sequence, majority sequence type, clade, and other variables:

```{r}
read_csv("../output/algae_df.csv") -> algae_df
```

- Checking data
```{r}
dim(algae_df)
head(algae_df)
```

- Examine the algal counts for two samples: 
```{r}
algae_df %>% 
  filter(sample_id == "PDOG-R1-1") %>% 
  arrange(-its2_count) %>% 
  group_by(Clade)


algae_df %>% 
  filter(sample_id == "SWAJ-R3-21") %>% 
  arrange(-its2_count, Clade) %>% 
  group_by(Clade)
```
- From the above, sample PDOG-R1-1 has the majority clade of A where the ITS2 abundance count is 21,483, and the rest of the ITS2 counts are 0. 

- For sample SWAJ-R3-21, the max ITS2 count is 20894, which is clade A. There are ITS2 counts for clade B, C, D, but we need the max clade per sample. 

- Need to obtain this assignment per sample before joining. 

- Group by sample id and clade, and then take the max its2 count. 

- Some samples have ITS2 counts for more than one clade.

- For the scope of this project, and since we need to join data from the ASV table, take only the majority clade. For ex, sample SYAN-R4-22 has ITS2 counts for clade A and C whereas most samples have only one clade. For this sample, take the clade with the higher ITS2 count, which in this case is clade A.
 
```{r}
algae_df %>% 
  group_by(sample_id, Clade)
```
 

```{r}
algae_df %>% 
  group_by(sample_id, Clade) %>% # yes, 659 samples
  # create new column of the majority clade's its2 counts (the max since rest are 0)
  mutate(max_clade_count = max(its2_count)) %>% 
  # ungroup data frame
  ungroup() %>% 
  # filter out the 0 values per sample
  filter(its2_count != 0) %>% # 778 rows, so some samples are giving extra rows
  filter(sample_id == "SYAN-R4-22")
```

- Filtering out the secondary clades as mentioned above: 
```{r}
algae_df %>% 
  group_by(sample_id, Clade) %>% # yes, 659 samples
  # create new column of the majority clade's its2 counts (the max since rest are 0)
  mutate(max_clade_count = max(its2_count)) %>% 
  # ungroup data frame
  ungroup() %>%  # 57,992  rows
  # filter out where the its2 count is 0
  filter(its2_count > 0) %>%  # 778 rows - therefore, some samples have two clades
  # arrange by the max_clade_count in order to slice per group
  arrange(-max_clade_count) %>% 
  # in order to slice the max clade count per group, need to group by sample id
  group_by(sample_id) %>%  # 658 samples, so somewhere, one of them has a 0 max count
  # based on arrangement above, this will grab max clade count row per sample if more than one. 
  dplyr::slice(1) %>% 
  # don't need this columns any longer as it's now the same as its2 count
  select(-max_clade_count) %>% 
  # reorder 
  select(sample_id, Clade, 
         ITS2, its2_count, 
         Majority, species, 
         region, reef) -> algae_clade_df


algae_clade_df %>% 
  # check that it sliced the right row for sample SYAN-R4-22
  filter(sample_id == "SYAN-R4-22")
  
```

```{r}
dim(algae_clade_df)
head(algae_clade_df)
```

- Write this to csv for later if needed
```{r}
write_csv(algae_clade_df, "../output/algae_clade_df.csv")
```

### Join Based on Sample ID

- Since there is one less sample in the algae df, use this with a left join to match up ASV columns for the 658 samples.
- Note: The ASV table was previously filtered down from ~33K to ~5,889 ASVs based on filtering criteria from Dr. Cárdenas. 

```{r}
algae_clade_df %>% 
  left_join(rev_bacteria, by = join_by(sample_id)) -> coral_data # joined algae and bacteria data
```

- Examine joined data
- Note that ITS2 should be a category not a numeric
```{r}
coral_data %>% 
  mutate(ITS2_f = as.factor(ITS2), .after = its2_count) %>% 
  select(-ITS2) -> coral_data

head(coral_data)
```

- Write to file
```{r}
write_csv(coral_data, "../output/coral_data.csv")
```





