---
title: "Metadata EDA"
author: "Cassandra Sperow"
date: "2023-09-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 

#### How many different regions? 6

#### How many different Lat/Long coordinates? 20

#### How many different reefs? 19

# Metadata EDA
```{r}
suppressMessages(library(tidyverse))
read_delim("../data/Files/metadata") -> metadata

metadata
```

- Dimensions of metadata table
```{r}
dim(metadata)
```
- Column names of metadata table
```{r}
names(metadata)
```

##### How many unique dates?  2 
```{r}
unique(metadata$Collection.date)
```

##### How many species? 2
```{r}
unique(metadata$Species)
```

##### How many Red Sea regions? 6 
```{r}
unique(metadata$Region)
```

##### How many different sample ids per reef? 19 reefs spread out amongst the 6 regions
```{r}
metadata %>% 
  group_by(Region, Reef) %>% 
  summarize(n = n())
```



##### How many unique pair of Lat/Long? 20 
```{r}
metadata %>% 
  group_by(Lat, Lon) %>% 
  summarise(n = n()) %>% 
  arrange(-n)
```

##### Map
- Where on a map are the samples taken from? 
```{r}
library(leaflet)
library(viridis)

palette_species <- colorFactor(viridis(2), metadata$Species)

leaflet() %>% 
  addProviderTiles(
    "Esri.WorldImagery",
    group = "Esri.WorldImagery"
  ) %>% 
  addMarkers(
    data = metadata,
    label = paste(
      "Sample: ", metadata$Sample, "; ",
      "Region: ", metadata$Region, "; ",
      "Species: ", metadata$Species
    )
  )
  # ) %>%
  # addCircleMarkers(
  #   data = metadata, 
  #   radius = 3, 
  #   fillOpacity = 1, 
  #   group = "Species"
  # )

```


```{r}
write.csv(metadata, "../output/metadata_map.csv")
```

