---
title: "PCA - Clade A"
author: "Cassandra Sperow"
date: "2023-11-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(library(tidyverse))
```

## Principal Components Analysis for Clade A
- Need to find the bacteria in order of those that explain the most variance for Clade A, and then do subsets of first 1000, 500. 
```{r, import-data-subset-A}
# read in data for clades A and C for Data 642 project
read_csv("../output/AC_df.csv") -> ac_df #  5889 ASVs

# section out clade A only
ac_df %>% 
  filter(Clade == "A") -> adf # clade A df
```


```{r, sum-to-zero}
# make sure no columns sum to 0
adf %>% 
  select(starts_with("ASV")) %>% 
  map(., ~sum(.)) %>% 
  as.data.frame() %>% 
  pivot_longer(cols = everything()) %>% 
  filter(value == 0) %>% 
### take these out of clade A df before PCA as they will cause an error
  select(name) %>% as_vector() -> zero_summed_clade_a

zero_summed_clade_a

adf %>% 
  select(-c(zero_summed_clade_a)) %>% 
  select(starts_with("ASV")) %>% 
  compositions::clr() %>% 
  as.data.frame() -> clr_asvs_clade_a
```


```{r, pca}
stats::prcomp(clr_asvs_clade_a, 
              scale. = T) -> pca_clade_a
```


```{r}
library(factoextra)
fviz_contrib(pca_clade_a, 
             choice = "var", 
             axes = 1, 
             top = 20, 
             title = "Clade A Top 20 Contributing ASVs in Dimension 1", 
             ggtheme = theme_minimal()) -> PCA_Clade_A_Top_20

PCA_Clade_A_Top_20
```


```{r}
fviz_contrib(pca_clade_a, 
             choice = "var", 
             axes = 1, 
             top = 400, 
             title = "Clade A Top Contributing ASVs in Dimension 1\nRed dashed line indicates average contribution", 
             xtickslab.rt = 0,
             #ylim(0.4, 0.7),
             ggtheme = theme_minimal())
```
- This is very different than Clade C. Clade C had a plateau of about 130-150 or so ASVs that contributed much more than the rest. 
- See the Rmd file for PCA_CLR that has other PCA analysis for CLade C towards the end of the file. 

```{r}
fviz_pca_biplot(pca_clade_a, 
                label = "var", 
                addEllipses=TRUE, 
                select.var = list(contrib = 5), 
                repel = T, 
                title = "Clade A - PCA Biplot"
                )
```


```{r}
fviz_pca_var(pca_clade_a, 
             col.var="contrib", 
             alpha.var="contrib", 
             select.var = list(contrib = 5), 
             repel = T,
             title = "Clade A Variables - PCA"
             ) 
```
```{r}
# Access the loadings of the first two principal components
loadings <- pca_clade_a$rotation[, 1:2]

# Display the loadings to see which variables contribute the most to PC1 and PC2
#loadings

# access ordered asv names for binding columns
names(clr_asvs_clade_a) -> asv_names

# Take the absolute values of the loadings
abs(loadings) %>% 
  as.data.frame()  -> abs_loadings

# need the above in particular format for map function
map(abs_loadings, ~abs(.)) %>% 
  as.data.frame() %>% 
  bind_cols("ASV" = asv_names, 
            .) %>% 
  arrange(-PC1) %>% 
  select(ASV) -> clade_a_pca_Ranks
```


```{r}
write_csv(clade_a_pca_Ranks, "../output/clade_a_pca_Ranks.csv")
```

# Subset PCA Ranked Variables for Clade A Dataset
```{r}
dim(adf) # clade A only
dim(ac_df) # Clades A and C together

# using df with both clades to be used later for classification for clade a = 1, clade c = 0
ac_df %>% 
  select(1:8, c(clade_a_pca_Ranks$ASV[1:1000])) %>% 
  # take out ITS2 type bc it is how to determine the clade
  # take out Majority bc it is also how clade is determined
  # take out reef bc it is not unique/specific enough on its own
  select(-ITS2_type, -Majority, -reef) -> pca_ranked_clades_AC_df
```


```{r}
#write_csv()
```


```{r}
```


```{r}
```

