---
title: "PCA_CLR_based_data"
author: "Cassandra Sperow"
date: "2023-10-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
suppressMessages(library(tidyverse))
```

## Read in Transformed Data for Principal Components Analysis
- Centered log ratio = CLR
- CLR data transformation recommended by client
```{r}
read_csv("../output/clr_bac_df.csv") -> clr_bac_df
```


```{r}
head(clr_bac_df) # not scaled, only clr transformation
```
## Run PCA - Both Species
- Scale = TRUE
```{r}
stats::prcomp(clr_bac_df, scale. = T) -> pca_out_clr

```

### Screeplot with base::plot()

```{r}
screeplot(pca_out_clr)
```

### Factoextra PCA CLR Visualization - Both Species
```{r}
library(factoextra)
fviz_pca_var(pca_out_clr, 
             #col.var = "steelblue", 
             col.var = "contrib", 
             gradient.cols = c("red", "blue", "green"), 
            # repel = T, 
            label = "none",
             title = "All Variables in PCA using CLR Transformation")
```


```{r}
summary(pca_out_clr)$importance %>% as_tibble() -> importance_matrix

```

##### Bind Columns of Names of PCA Outputs to View Percentages
```{r}
bind_cols(
  name = c("StndDev", "IndivProp", "CumProp"), 
  importance_matrix
) -> importance_df
```

#### Screeplot vis Ggplot
```{r}
importance_df %>% 
  t() %>% 
  as.data.frame() %>% 
  janitor::row_to_names(1) %>% 
  ggplot(aes(x = as.numeric(CumProp) , y =  as.numeric(IndivProp))) +
  geom_point() +
  ggtitle("PCA Results: Proportion of Variance", "This is a better screeplot showing all PCs on CLR-based scaled data") +
  xlab("Cumulative Proportion") +
  ylab("Proportion of Variance per Principal Component") +
  theme_bw()
```


### How many PCs contribute at least 80 % ? 361
```{r}
importance_df %>% 
  # pivoting data drame so that cumulatives are in column
  t() %>%  as.data.frame() %>% 
  # moving up top row as column header
  janitor::row_to_names(1) %>% ######## snapshot here for presentation
  # filtering cumulative proportion column for greater > = 80 %
  filter(CumProp >= 0.8 & CumProp <= 0.78) 

```



# Split by Species for PCA


## Species P - Import Data

```{r}
# add back in the sample IDs to separate by species
# read in original data frame from creating the CLR transformation
read_csv("../output/rev_bacteria.csv") -> rev_bacteria

# section out the sample ID column and save
rev_bacteria[,1] -> sample_id_column

# re-combine with clr-transformed data and save
bind_cols(
  sample_id_column,
  clr_bac_df
) -> clr_bac_df_samples

```

#### Split for Species P
```{r}
clr_bac_df_samples %>% 
  # the sample IDs that have a capital S at the beginning of the string are species S
  filter(str_detect(sample_id, "^P")) -> species_P_CLR_DF
```

- PCA can't rescale a constant 0 column
- Check the columns that sum to 0:  1,110 columns sum to 0 and would not allow PCA to work for species P
```{r}
# PCA can't rescale a constant 0 column
# Check transformed data for columns that are all 0 values
species_P_CLR_DF[,-1] %>% 
  map(., ~sum(.)) %>% 
  as.data.frame() %>% 
  pivot_longer(cols = everything()) %>% 
  filter(value == 0) -> zero_columns

```

- Remove columns identified above for successful run of PCA
```{r}
species_P_CLR_DF %>% 
  select(-(zero_columns$name)) -> pca_data_species_P
```


### Species P - PCA


```{r}
stats::prcomp(pca_data_species_P[,-1], scale. = T) -> pca_out_P
```


```{r}
base::plot(pca_out_P)
```


```{r}
library(factoextra)
fviz_pca_var(pca_out_P, 
             #col.var = "steelblue", 
             col.var = "contrib", 
             gradient.cols = c("red", "blue", "green"), 
             #repel = T, 
            label = "none",
             title = "ASV Variables in Species P")
```

#### Which of these explain at least 80 % of the variability in species P?
```{r}
summary(pca_out_P)$importance %>% as_tibble() -> importance_matrix_P
```


```{r}
bind_cols(
  name = c("StndDev", "IndivProp", "CumProp"), 
  importance_matrix_P
) -> species_p_importance
```


```{r}
species_p_importance %>% 
  t() %>% 
  as.data.frame() %>% 
  janitor::row_to_names(1) %>% 
  ggplot(aes(x = as.numeric(CumProp) , y =  as.numeric(IndivProp))) +
  geom_point() +
  ggtitle("Species P - PCA Results: Proportion of Variance") +
  xlab("Cumulative Proportion") +
  ylab("Proportion of Variance per Principal Component") +
  theme_bw()
```

##### How many PCs contribute to at least 80 % ? 166
```{r}
species_p_importance %>% 
  # pivoting data drame so that cumulatives are in column
  t() %>%  as.data.frame() %>% 
  # moving up top row as column header
  janitor::row_to_names(1) %>% ######## snapshot here for presentation
  # filtering cumulative proportion column for greater > = 80 %
  filter(CumProp >= 0.8)
```


```{r}
fviz_pca_ind(pca_out_P, 
             geom.var = c("point", "text"),
            # repel = T,
             title = "Species P - ASV Individuals")


fviz_pca(pca_out_P, 
        # col.var = "contrib", 
         #select.ind = list(contrib = 10)
         select.var = list(contrib = 2),
         #repel = T
        #label = "none"
         )
```

### Get PCA Variable Names
```{r}
get_pca_var(pca_out_P)
```


```{r}
get_pca_var(pca_out_P)$contrib %>% 
  as.data.frame() %>% 
  arrange(-Dim.1) -> species_P_pca
```

```{r}
write_csv(species_P_pca, "../output/species_P_pca_ranks.csv")
```


```{r}
fviz_contrib(pca_out_P, choice = "var", axes = 1, top = 20, 
             title = "Species P Top 20 Contributing ASVs in Dimension 1", 
             ggtheme = theme_minimal()) -> top_20_pca_P

top_20_pca_P
```

```{r}
ggsave("../plots/top_20_pca_P.png", top_20_pca_P, width = 14, bg = "#ffffffff")
```

# Species S
## Species S - Import Data

```{r}
# section out the observations for species S and save
clr_bac_df_samples %>% 
  filter(str_detect(sample_id, "^S")) -> clr_data_S

```


- PCA can't rescale a constant 0 column
- Check the columns that sum to 0 for species S:
```{r}
# PCA can't rescale a constant 0 column
# Check transformed data for columns that are all 0 values
clr_data_S[,-1] %>% 
  map(., ~sum(.)) %>% 
  as.data.frame() %>% 
  pivot_longer(cols = everything()) %>% 
  filter(value == 0) -> zero_columns_S

## take the columns that sum to 0 out before running PCA for species S
clr_data_S %>% 
  select(-(zero_columns_S$name)) -> pca_data_species_S

```


## Run PCA for Species S
```{r}
stats::prcomp(pca_data_species_S[,-1], # take out sample_id column
              scale. = T) -> pca_out_S
```

```{r}
fviz_pca_var(pca_out_S, 
             #col.var = "steelblue", 
             col.var = "contrib", 
             gradient.cols = c("red", "blue", "green"), 
             #repel = T, 
            label = "none",
             title = "ASV Variables in Species S")
```
- This is quite different from Species P where the highest contributing only went into one quadrant. See above. 

```{r}
get_pca_var(pca_out_S)
```


```{r}
get_pca_var(pca_out_S)$contrib %>%
  as.data.frame() %>% 
  arrange(-Dim.1) -> species_S_pca_ranks
```

```{r}
write_csv(species_S_pca_ranks, "../output/species_S_pca_ranks.csv")
```


```{r}
fviz_contrib(pca_out_S, choice = "var", axes = 1, top = 20, 
             title = "Species S Top 20 Contributing ASVs in Dimension 1", 
             ggtheme = theme_minimal())  -> top_20_pca_S

top_20_pca_S
```


```{r}
ggsave("../plots/top_20_pca_S.png", top_20_pca_S, width = 14, bg='#ffffff')
```

